Fisher vector places: learning compact descriptors for place recognition

### 2014-11-13 cvpr15review.tex
edited blue text

\paragraph{Large-scale visual place recognition.} %\vspace{-3mm}
    The visual localization problem is typically treated as a large-scale instance-level retrieval~\cite{Cummins09,Chen11,Gronat13,Knopp2010,Schindler07,Torii2013,Zamir10}, where images are represented using local invariant features~\cite{Lowe04} aggregated into the bag-of-visual-words~\cite{Csurka04,Sivic03} representation. The image database can be further augmented by 3D point clouds~\cite{Klinger13}, automatically reconstructed by large-scale structure from motion (SfM)~\cite{Agarwal-ICCV-2009,Klinger13}, which enables accurate prediction of query image camera position~\cite{Li12,Sattler12}. In this work we investigate learning a discriminative representation using the compact Fisher vector descriptors~\cite{Jegou12}. Fisher vector descriptors have shown excellent place recognition accuracy~\cite{Torii2013}. In this work we further improve their performance by discriminative learning.

@@

\section{Learning compact place descriptors using per-exemplar SVM}
\label{sec:perExemplar}
   Each database image $j$ is represented by its L2-normalized Fisher vector $\Phi_j$. The goal is to learn a set of new L2-normalized Fisher vectors $\Psi_j$, one per each database image.
   % such that at query time, 
   \textcolor{petr}{At query time,} given the Fisher vector $\Phi_q$ of an unknown query image, we retrieve the database image depicting the same location by finding the image $j^*$ with the highest score measured by a dot product
   %
   \begin{equation}
       j^*=\operatorname*{arg\;max}_{j} \Phi_q^T \Psi_j. 
       \label{eq:class}
   \end{equation}
   %  
   In other words, the aim is to replace each original database Fisher vector $\Phi_j$ with a new vector $\Psi_j$ that is more discriminative in the sense of separation from descriptors of images depicting other places. Inspired by~\cite{Gronat13}, we investigate applying the exemplar support vector machine (e-SVM)~\cite{Malisiewicz11} for this task. e-SVM learns a linear classifier $w_j^T\Phi+b_j$ given the descriptor $\Phi_j^+$ of place $j$ as a single positive example (with target label +1) and a large number of negative descriptors $\mathcal N_j$ from other places in the database (with target labels -1). The intuition of the exemplar SVM training~\cite{Malisiewicz11} is that the learnt weight vector $w_j$ will give a higher weight to the dimensions of the descriptor that are discriminative and will down-weight dimensions that are non-discriminative with respect to the negative training data collected from other far-away places.  The optimal $w_j$ and $b_j$ are obtained by minimizing the following objective  
      %
      \begin{align}
         \label{eq:obj}
         ||w_j||^{2} &+C_1 \cdot h
            \left(
               w_j^T\Phi^+_j+b_j
            \right)  \\ \nonumber
            &+C_2\sum_{\Phi\in \mathcal N_j}h
            \left(
               -w_j^T\Phi-b_j
            \right),    
      \end{align}
      %
   where $\Phi^+_j$ is the descriptor of the place $j$ as the positive data point, $\Phi$ are Fisher descriptors from negative training data $\mathcal N_j$ and $h$ is the hinge loss, $h(y) = \max(1-y,0)$. Note that the first term in~(\ref{eq:obj}) is the regularizer, the second term is the loss on the positive data weighted by scalar parameter $C_1$ and the third term is the loss on the negative data weighted by scalar parameter $C_2$. The objective is convex and can be minimized with respect to $w_j$ and $b_j$ using standard software packages such as~\cite{libsvm}. A key advantage is that the per-exemplar classifier for each place can be trained independently and hence the learning can be heavily parallelized. The downside of the independent training for each positive example is that the resulting scores have to be calibrated with respect to each other on additional data~\cite{Gronat13,Malisiewicz11}.

@@

   \subsection{Interpreting normalized $w$ as a new descriptor} %\vspace{-0.2cm} 
      The above analysis demonstrates that as $C_2$ decreases the normal of the optimal hyperplane $w$ that separates the positive exemplar $\Phi^+$ from negative data becomes parallel with $\Phi^+$, as shown in figure \ref{fig:C2effect}. As $C_2$ increases, the normal $w$ of the optimal hyper-plane moves away from the direction given by the positive example $\Phi^+$ in a manner that reduces the loss on the negative data. 
      This suggests that the learnt $w$ could be interpreted as a modified positive example $\Phi^+$, re-weighted to emphasize directions that separate $\Phi^+$ from the negative data. As discussed above $w$ is not normalized. As we wish to measure the similarity between descriptors by (the cosine of) their angle given by equation \eqref{eq:class}, additional normalization of the learnt $w$ is necessary. Hence we define the new descriptor $\Psi_j$ as the normalized hyperplane normal $w_j$
         %
         \begin{equation}
            \label{eq:normalization}
            \Psi_j=\dfrac{w_j}{||w_j||}.
         \end{equation}

      \textcolor{petr}{%TODO: \emph{we should't speak about calibration, we shoud address not ignoring $b$}\\
         The intuition is that for L2 normalized query descriptors $\Phi_q$ computing the calibrated score $f(x)$ corresponds to computing the normalized cross-correlation between vectors $\Psi_j$ and $\Phi_q$. This was found to work well in retrieval (see e.g.~\cite{Sivic03}) or matching whitened HOG descriptors \cite{Gharbi12, Hariharan12}, but has not yet been used in conjunction with SVMs. 
         \\
         Notice that even the new descriptor $\Psi_j$ is obtained by normalization of the learnt hyperplane $w_j$, the bias $b$ form equation~\eqref{eq:decomposed} is not ignored. Indeed, the bias term in convex objective~\eqref{eq:obj} is affecting the learnt the hyperplane.
         %\\
         %The proposed re-normalization procedure can be interpreted as a simple form of affine calibration where the uncalibrated SVM score $s(x)=w^T x+b$ is transformed into a calibrated score $f(x)=\alpha s(x)+\beta$, where $\alpha = 1/||w||$ and $\beta = -b\alpha$. Note that the bias term b is not ignored but used to compute $f(x)$. The intuition is that for L2 normalized descriptors x computing the calibrated score $f(x)$ corresponds to computing the normalized cross-correlation between vectors w and x. This was found to work well in retrieval (see e.g. Sivic and Zisserman, ICCV 2003 [32]) or matching whitened HOG descriptors [11,13] (see e.g. [Doersch et al., Mid-Level Visual Element Discovery as discriminative Mode Seeking, NIPS 2013]), but has not yet been used in conjunction with SVMs as done in this work. Note also that the affine calibration can be computed offline and does not need any additional computation or storage at query time compared to the p-value calibration of Gronat et al. [12].
      }

@@

   \subsection{Results}  
      For each database (Pittsburgh 25k and Pittsburgh 55k) we compare results of our method (FV e-SVM) to two baselines: standard bag-of-visual-words baseline (BOW) and raw Fisher vector matching without learning (FV).

@@

      \paragraph{Applying our learning method to other descriptors.}
        \textcolor{petr}{ 
         The proposed approach can be applied to other descriptors beyond Fisher vectors. To demonstrate this we have applied the proposed method to the bag-of-visual-words descriptor and observed improvement from 28.7\% recall@K=1 (vanilla bag-of-visual-words) to 31.8\% recall@K=1 (e-SVM). For bag-of-visual-words, the e-SVM with p-value calibration gives comparable results (33.6\% recall@K=1) but at an increased training cost and a significantly higher required memory footprint at query time, which makes the method very hard to scale-up to database sizes beyond 200k images. The results are given in table \ref{tab:recall}.
        }

@@

         On the Pittsburgh 25k database, we compare performance of our learnt discriminative descriptors to the method of~\cite{Gronat13} who report on the similar testing data. Our method outperforms~\cite{Gronat13} already for dimension $d=128$ (37.8\%).
         \\
         The p-val calibration proposed by Gronat et al.~\cite{Gronat13} was also tested on e-SVM Fisher vectors. We found that the calibration did not perform for the learnt Fisher vector classifiers (top 1 recall of 25.3\% compared to baseline performance of 33.6\% for dimension 128). When examining the results, we believe this may be attributed to the fact that Fisher vectors are low dimensional compared to the extremely high-dimensional (d=100,000) bag-of-visual-words representation, which affects the distribution of the classifier scores on the negative test data.
         \\
         Finally, we investigate how performs a raw exemplar-SVM without any further calibration both image descriptors . We observe that while using raw SVM scores directly without any calibration fails on BOW (as reported in \cite{Gronat13}), it marginally improves the recall for larger $K$ for Fisher vectors as indicated in \ref{tab:recall}.

@@

                     in top $K\in\{1,2,5,10,20\}$ retrieved images.
                     Different methods have been applied to two types of descriptors, the BOW and Fisher vectors compressed to different dimensions.The learnt representations (BOW \emph{e-SVM} and BOW \emph{p-val}) outperform the raw BOW baseline as well as the learnt representation without calibration (BOW raw SVM).

@@

                  The learnt descriptors by the proposed method (FV e-SVM) consistently improve over the raw Fisher vector descriptors across the whole range of $K$  and all dimensions on both the 25k and 55k Pittsburgh image datasets.     

@@

         \begin{table}[t]
            \begin{centering}
               \include{table55k}
               \caption{ \textbf{Results on Pittsburgh 55k dataset.}
                  The fraction of correctly recognized queries (recall@K)
                  \textcolor{petr}{
                  in top $K\in\{1,2,5,10,20\}$ retrieved images.
                  The proposed method (\emph{e-SVM}) has been applied to Fisher vectors compressed to different dimensions.
                  } 
                  The learnt descriptors by the proposed method (FV e-SVM) consistently improve over the raw Fisher vector descriptors across the whole range of $K$  and all dimensions on both the 25k and 55k Pittsburgh image datasets.     
               }
            \label{tab:recall2}
            \end{centering}
         \end{table}
         %
         %
         %
         \begin{figure*}[h]
            \include{myFigureWnorm}
            \caption{
                \textbf{Examples of correctly and incorrectly localized queries for the learnt Fisher vector representation.}
                Each example shows a query image (left) together with correct (green) and incorrect (red) matches from the database obtained by the learnt Fisher vector representation \emph{e-SVM} method (top) and the standard Fisher vector baseline (bottom) for dimension 128. Note that the proposed method is able to recognize the place depicted in the query image despite changes in viewpoint, illumination and partial occlusion by other objects (trees, lamps) and buildings. 
                Note that the baseline methods often finds images depicting the same buildings but in a distance whereas our learnt representation often finds a closer view better matching the content of the query.       
            }
            \label{fig:images}
         \end{figure*}
         %
         %
         %

>>>>

\paragraph{Large-scale visual place recognition.} %\vspace{-3mm}
    The visual localization problem is typically treated as a large-scale instance-level retrieval~\cite{Cummins09,Chen11,Gronat13,Knopp2010,Schindler07,Torii2013,Zamir10}, where images are represented using local invariant features~\cite{Lowe04} aggregated into the bag-of-visual-words~\cite{Csurka04,Sivic03} representation. The image database can be further augmented by 3D point clouds~\cite{Klinger13}, automatically reconstructed by large-scale structure from motion (SfM)~\cite{Agarwal-ICCV-2009,Klinger13}, which enables accurate prediction of query image camera position~\cite{Li12,Sattler12}. 
     In this work we investigate learning a discriminative place-specific image representation. A similar idea has been recently explored in~\cite{Cao13} 
    who learn a graph-based discriminative representation for landmark image collections where typically many images are available for each place.
    In contrast, we focus on street-level images such as Google street-view, which have greater coverage, but typically only one or a small number of images are  available for each place.  
     We investigate learning a discriminative representation using the compact Fisher vector descriptors~\cite{Jegou12}. Fisher vector descriptors have shown excellent place recognition accuracy~\cite{Torii2013}. In this work we further improve their performance by discriminative learning.

@@

\section{Learning compact place descriptors using per-exemplar SVM}
\label{sec:perExemplar}
   Each database image $j$ is represented by its L2-normalized Fisher vector $\Phi_j$. The goal is to learn a set of new L2-normalized Fisher vectors $\Psi_j$, one per each database image.
   % such that at query time, 
   At query time, given the Fisher vector $\Phi_q$ of an unknown query image, we retrieve the database image depicting the same location by finding the image $j^*$ with the highest score measured by a dot product
   %
   \begin{equation}
       j^*=\operatorname*{arg\;max}_{j} \Phi_q^T \Psi_j. 
       \label{eq:class}
   \end{equation}
   %  
   In other words, the aim is to replace each original database Fisher vector $\Phi_j$ with a new vector $\Psi_j$ that is more discriminative in the sense of separation from descriptors of images depicting other places. Inspired by~\cite{Gronat13}, we investigate applying the exemplar support vector machine (e-SVM)~\cite{Malisiewicz11} for this task. e-SVM learns a linear classifier $w_j^T\Phi+b_j$ given the descriptor $\Phi_j^+$ of place $j$ as a single positive example (with target label +1) and a large number of negative descriptors $\mathcal N_j$ from other places in the database (with target labels -1). The intuition of the exemplar SVM training~\cite{Malisiewicz11} is that the learnt weight vector $w_j$ will give a higher weight to the dimensions of the descriptor that are discriminative and will down-weight dimensions that are non-discriminative with respect to the negative training data collected from other far-away places.  The optimal $w_j$ and $b_j$ are obtained by minimizing the following objective  
      %
      \begin{align}
         \label{eq:obj}
         ||w_j||^{2} &+C_1 \cdot h
            \left(
               w_j^T\Phi^+_j+b_j
            \right)  \\ \nonumber
            &+C_2\sum_{\Phi\in \mathcal N_j}h
            \left(
               -w_j^T\Phi-b_j
            \right),    
      \end{align}
      %
   where $\Phi^+_j$ is the descriptor of the place $j$ as the positive data point, $\Phi$ are Fisher descriptors from negative training data $\mathcal N_j$ and $h$ is the hinge loss, $h(y) = \max(1-y,0)$. Note that the first term in~(\ref{eq:obj}) is the regularizer, the second term is the loss on the positive data weighted by scalar parameter $C_1$ and the third term is the loss on the negative data weighted by scalar parameter $C_2$. The objective is convex and can be minimized with respect to $w_j$ and $b_j$ using standard software packages such as~\cite{libsvm}. A key advantage is that the per-exemplar classifier for each place can be trained independently and hence the learning can be heavily parallelized. The downside of the independent training for each positive example is that the resulting scores have to be calibrated with respect to each other on additional data~\cite{Gronat13,Malisiewicz11}.

@@

   \subsection{Interpreting normalized $w$ as a new descriptor} %\vspace{-0.2cm} 
      The above analysis demonstrates that as $C_2$ decreases the normal of the optimal hyperplane $w$ that separates the positive exemplar $\Phi^+$ from negative data becomes parallel with $\Phi^+$, as shown in figure \ref{fig:C2effect}. As $C_2$ increases, the normal $w$ of the optimal hyper-plane moves away from the direction given by the positive example $\Phi^+$ in a manner that reduces the loss on the negative data. 
      This suggests that the learnt $w$ could be interpreted as a modified positive example $\Phi^+$, re-weighted to emphasize directions that separate $\Phi^+$ from the negative data. As discussed above $w$ is not normalized. As we wish to measure the similarity between descriptors by (the cosine of) their angle given by equation \eqref{eq:class}, additional normalization of the learnt $w$ is necessary. Hence we define the new descriptor $\Psi_j$ as the normalized hyperplane normal $w_j$
         %
         \begin{equation}
            \label{eq:normalization}
            \Psi_j=\dfrac{w_j}{||w_j||}.
         \end{equation}
      \textcolor{petr}{%TODO: \emph{we should't speak about calibration, we shoud address not ignoring $b$}\\
%      Measuring the similarity between descriptors by their normalized cross-correlation 
         Note again that for L2 normalized query descriptor $\Phi_q$, evaluating $\Phi_q^T\Psi_j$ in eq.~\eqref{eq:class} corresponds to computing the normalized cross-correlation between $\Phi_q$ and $\Psi_j$. This was found to work well in retrieval (see e.g.~\cite{Sivic03}) or matching whitened HOG descriptors \cite{Gharbi12, Hariharan12}, but has not yet been used in conjunction with SVMs.    
         Note that while the bias term $b$ is not directly used in the construction of $\Psi_j$, the bias affects the learnt optimal $w_j$ as it appears in the SVM objective~\eqref{eq:obj}. 
         % when learning the optimal SVM hyper-plane $w_j$.           
%         even the new descriptor $\Psi_j$ is obtained by normalization of the learnt hyperplane $w_j$, the bias $b$ form equation~\eqref{eq:decomposed} is not ignored. Indeed, the bias term in convex objective~\eqref{eq:obj} is affecting the learnt the hyperplane.
%         %\\
         %The proposed re-normalization procedure can be interpreted as a simple form of affine calibration where the uncalibrated SVM score $s(x)=w^T x+b$ is transformed into a calibrated score $f(x)=\alpha s(x)+\beta$, where $\alpha = 1/||w||$ and $\beta = -b\alpha$. Note that the bias term b is not ignored but used to compute $f(x)$. The intuition is that for L2 normalized descriptors x computing the calibrated score $f(x)$ corresponds to computing the normalized cross-correlation between vectors w and x. This was found to work well in retrieval (see e.g. Sivic and Zisserman, ICCV 2003 [32]) or matching whitened HOG descriptors [11,13] (see e.g. [Doersch et al., Mid-Level Visual Element Discovery as discriminative Mode Seeking, NIPS 2013]), but has not yet been used in conjunction with SVMs as done in this work. Note also that the affine calibration can be computed offline and does not need any additional computation or storage at query time compared to the p-value calibration of Gronat et al. [12].
      }

@@

   \subsection{Results}  
      For both databases (Pittsburgh 25k and Pittsburgh 55k) we compare results of our method (FV e-SVM) to two baselines: standard bag-of-visual-words baseline (BOW) and raw Fisher vector matching without learning (FV).  
%  For the Pittsburgh 25k database we also compare results to exemplar SVM learning without the proposed descriptor re-normalization (SVM raw).

@@

      \paragraph{Applying our learning method to other descriptors.}
        \textcolor{petr}{ 
         The proposed approach can be applied to other descriptors beyond Fisher vectors. To demonstrate this we have applied the proposed method to the bag-of-visual-words descriptor and observed improvement from 28.7\% recall@K=1 (BOW) to 31.8\% recall@K=1 (BOW e-SVM).  
For bag-of-visual-words,  the e-SVM with p-value calibration of~\cite{Gronat13} (BOW p-val) gives a slightly better results (33.6\% recall@K=1) but at an increased training cost and a significantly higher required memory footprint at query time, which makes the method hard to scale-up to database sizes beyond 200k images. The results are given in table \ref{tab:recall} (top four rows).
        }

@@

         On the Pittsburgh 25k database, we compare performance of our learnt discriminative descriptors to the method of~\cite{Gronat13} who report results on similar testing data. Our method outperforms~\cite{Gronat13} already for dimension $d=128$ (37.8\%).
  %       \\
         We have also tested the p-val calibration of~\cite{Gronat13} on e-SVM Fisher vectors. We found that the calibration did not perform well for the learnt Fisher vector classifiers (top 1 recall of 25.3\% compared to baseline performance of 33.6\% for dimension 128). When examining the results, we believe this may be attributed to the fact that Fisher vectors are low dimensional compared to the extremely high-dimensional (d=100,000) bag-of-visual-words representation, which affects the distribution of the classifier scores on the negative test data.
%         \\
         Finally, we investigate the performance of a raw exemplar-SVM (raw e-SVM) without any further calibration for both image descriptors . We observe that while using raw SVM scores directly without any calibration fails for the bag-of-visual-words representation (BOW raw SVM) (as reported in \cite{Gronat13}), it marginally improves recall for larger $K$ for Fisher vectors (68.4\% for FV128 raw SVM compared to 67.7\% for non-learned FV128).
%         as indicated in \ref{tab:recall} (FV128 raw SVM).

@@

                     in the top $K\in\{1,2,5,10,20\}$ retrieved images.
%                     Different methods have been applied to two types of descriptors, the BOW and Fisher vectors compressed to different dimensions. The learnt representations (BOW \emph{e-SVM} and BOW \emph{p-val}) outperform the raw BOW baseline as well as the learnt representation without calibration (BOW raw SVM).

@@

                  The learnt descriptors by the proposed method (FV e-SVM) consistently improve over the raw Fisher vector descriptors (FV) across the whole range of $K$  and all dimensions.
%                   on both the 25k and 55k Pittsburgh image datasets.     

@@

         \begin{table}[t]
            \begin{centering}
               \include{table55k}
               \caption{ \textbf{Results on Pittsburgh 55k dataset.}
                  The fraction of correctly recognized queries (recall@K)
                  \textcolor{petr}{
                  in top $K\in\{1,2,5,10,20\}$ retrieved images.
                  The proposed method (\emph{e-SVM}) has been applied to Fisher vectors compressed to different dimensions.
                  } 
                  The learnt descriptors by the proposed method (FV e-SVM) consistently improve over the raw Fisher vector descriptors across the whole range of $K$  and all dimensions.
%                   on both the 25k and 55k Pittsburgh image datasets.     
               }
            \label{tab:recall2}
            \end{centering}
         \end{table}
         %
         %
         %
         \begin{figure*}[h]
            \include{myFigureWnorm}
            \caption{
                \textbf{Examples of correctly and incorrectly localized queries for the learnt Fisher vector representation.}
                Each example shows a query image (left) together with correct (green) and incorrect (red) matches from the database obtained by the learnt Fisher vector representation \emph{e-SVM} method (top) and the standard Fisher vector baseline (bottom) for dimension 128. Note that the proposed method is able to recognize the place depicted in the query image despite changes in viewpoint, illumination and partial occlusion by other objects (trees, lamps) and buildings. 
                Note that the baseline methods often finds images depicting the same buildings but in a distance whereas our learnt representation often finds a closer view better matching the content of the query.       
            }
            \label{fig:images}
         \end{figure*}
         %
         %
         %
